Fast radio burst cosmology: dispersion measure angular power spectrum

### 2015-02-26 writeup/dm_angular_ps.tex
dm_space_distortions

In this \emph{ Letter} we consider the implications of of a cosmological population of fast radio bursts (FRBs). 
If FRBs originate at cosmological distances they provide a new window on the Universe. The key insight into the potential power of FRBs is that, while it is interesting and helpful to know the redshift of the burst, it is not necessary to do interesting studies of cosmology.

@@

In addition, we define the mean subtracted values of the of the variouse dispersion
measure quantites, $\del D = D - \langle D \rangle$. Here the angular brackets represent
the ensemble average.

@@

\red{Convensions: we use subscripts $i$, $j$, etc.~on coordinate-like
    quantities to label observed FRB events. Thus $\vec x_i$ and $\hat{n}_i$
    are respectively the 3D coordinates of and unit vector pointing to event $i$.
    This should not be confused with superscripts, denoting the
component of a vector, such that $x^1$ is the first componenent of vector $\vec
x$.  }

@@

@@

The principle observable which we would like to consider is the correlations
between the dispersion measures of observed fast radio bursts as a function of
their angular separation.  For observed FRB events labeled by the indices $i$
and $j$, and assuming $i \neq j$, we define the dispersion measure angular
correlation function as
\begin{equation}
    w(\hat{n}_i\cdot\hat{n}_j) \equiv \langle \del D_i \del D_j \rangle.
\end{equation}
A move convienient quantity to work with is the corresponding angular
power spectrum
\begin{equation}
w(\hat{n}_i \cdot\hat{n}_j) = \sum_l \frac{2l+1}{4 \pi} C_l
P_l(\hat{n}_i\cdot\hat{n}_j).
\end{equation}

We have identified two effects that dominate the angular
correlations and which depend on the large-scale structure of the Universe and
thus cosmology.  The first effect is that the signals from two FRB events may
propagate through shared structures along the line of sight.  Thus correlations
in the election overdensities along the line of sight will correlate the 
$\Dde$.  The second effect is that due to the intrisic clustering of source
events, a pair of FRBs that are nearby in the angular coordinates are more
likely to also be nearby in the line of sight coordinate, thus causing
correations in $\Dne$.  In the remainder of this section we will calculate the
contributions from each of these effects.

We assume that $D^{\rm src}_i$ are uncorrelated. Again this may not
be strictly true due to clustering of FRB sources and environmental effects
and this may bias results.  In addition $\Dsrc$ and $\Dne$ may be correlated due
to redshift evolution of source events.

@@

@@

The joint probablility of observing a pair of FRB events at 3D coordinates
$\vec x_i$ and $\vec x_j$ in infintessimal volumes $\ud^3\vec x_i$ and
$\ud^3\vec x_j$ is
\begin{equation}
    \calP(\vec x_i, \vec x_j) = p(\eta_i) p(\eta_j) \left[ 1 + b(z_i)
    b(z_j) \xi(\vec x_i - \vec x_j) \right] \frac{\ud^3\vec x_i}{\eta_i^2}
    \frac{\ud^3\vec x_j}{\eta_j^2}.
\end{equation}
The factors of $1/\eta^2$ are nessisary to achive the proper volume weighting,
a fact that can be seen by taking the limit of $\xi \to 0$ and integrating of
the angular coordinates. Noting that $p(\eta_i)$ has units 
\red{$({\rm MPc}\,{\rm rad}^2)^{-1}$ (conflicts with being normalized)}.

An observed event has known angular coordinates, and we are interested in
correlations in $\Dne$ which is directly related to the line of sight
coordinate.  Fixing the angular coordinates:
\begin{equation}
    \calP(\eta_i, \eta_j | \hat{n}_i, \hat{n}_i) = N(\hat{n}_i, \hat{n}_i)
    p(\eta_i)p(\eta_j) \left[ 1 + b(z_i)
    b(z_j) \xi(\vec x_i - \vec x_j) \right] \ud \eta_i \ud \eta_j,
\end{equation}
where $N(\hat{n}_i, \hat{n}_i)$ is a normalizing factor which ensures the
expression integrates to unity over $\eta_i$ and $\eta_j$. The factors of
$1/\eta^2$ have been canceled when converting $\ud^3 \vec x = \eta^2 \ud \eta \ud^2
\vec \theta$. If the second term in brackets is small, then the normalizations
$N$ is 1 (since the distributions $p{\eta_i}$ are already normalized. 
This expression is
only valid if $i \ne j$.
Transforming to $\Dne$ as our distance measure, we obtain
\begin{align}
\calP(\Dne_i, \Dne_j | \hat{n}_i, \hat{n}_i)
    &= p(\eta_i)p(\eta_j) \left[ 1 + b(z_i)
        b(z_j) \xi(\vec x_i - \vec x_j) \right]
        \frac{\ud \eta_i}{\ud \Dne_i} \frac{\ud \eta_j}{\ud \Dne_j}
        \ud \Dne_i\ud \Dne_j
        \\
    &= p(\eta_i)p(\eta_j) \left[ 1 + b(z_i)
        b(z_j) \xi(\vec x_i - \vec x_j) \right]
        \frac{\ud \Dne_i}{1 + z_i}\frac{\ud \Dne_j}{1 + z_j}
\end{align}
Finally, taking the second moment and subtracting off the product of the first
moments:
\begin{equation}
\langle \Dne_i \Dne_j \rangle
    - \langle \Dne_i \rangle \langle \Dne_j \rangle =
    \int \Dne_i \Dne_j
    \frac{p(\eta_i)}{1 + z_i}\frac{p(\eta_j)}{1 + z_j}
    b(z_i) b(z_j) \xi(\vec x_i - \vec x_j)
        \ud \Dne_i\ud \Dne_j
\end{equation}
And thus we have:
\begin{equation}
    w^{\bar{n}_e \bar{n}_e}(\hat{n}_i\cdot\hat{n}_j)
        =\int \Dne_i \Dne_j
    \frac{p(\eta_i)}{1 + z_i}\frac{p(\eta_j)}{1 + z_j}
    b(z_i) b(z_j) \xi(\eta_i \hat{n}_i - \eta_j \hat{n}_j)
        \ud \Dne_i\ud \Dne_j
\end{equation}
Where $\eta$ and $z$ are understood to be functions of $\Dne$.

@@

Here we sketch how the dispersion mearure angular power-spectrum, $C_l$
could be estimated from a sample of observed dispersion-measures.

@@

We see that the off diagonal terms depend on the angular power-spectrum of
interest while the diagonal is contaminiated by the unknown parameter
$\sigma_{\rm src}^2$. This is again in analogue to weak lensing where the
correlations between galaxy elipticty are caused primarily by cosmic leansing shear
while an individual galaxies elipticity is dominated the galaxy's intrinsic
elipticity. To remove the information in the diagonal, we define a
modified covariance matrix derivative such that:
\begin{equation}
    \tilde{C}_{ij,\alpha} \equiv
\begin{cases}
    C_{ij,\alpha} & \text{if } i \neq j,\\
0 & \text{if } i = j
\end{cases}.
\end{equation}

>>>>

In this \emph{Letter} we consider the implications of of a cosmological population of fast radio bursts (FRBs). 
If FRBs originate at cosmological distances they provide a new window on the Universe. The key insight into the potential power of FRBs is that, while it is interesting and helpful to know the redshift of the burst, it is not necessary to do interesting studies of cosmology.

@@

In addition, we define the mean subtracted values of the of the various dispersion
measure quantities, $\del D = D - \langle D \rangle$. Here the angular brackets represent
the ensemble average.

@@

\red{Conventions: we use subscripts $i$, $j$, etc.~on coordinate-like
    quantities to label observed FRB events. Thus $\vec x_i$ and $\hat{n}_i$
    are respectively the 3D coordinates of and unit vector pointing to event $i$.
    This should not be confused with superscripts, denoting the
component of a vector, such that $x^1$ is the first component of vector $\vec
x$.  }

@@

An important characteristic for an FRB survey is the ensemble radial distribution
function $p(\eta_i)$ which gives the probability density that an FRB source event 
occurred at radial distance $\eta_i$. The number events between radial distance
$\eta$ and $\eta + \ud \eta$ is
$N p(\eta)$ where $N$ is the cumulative number of events in the sample. The 3D source
event density is $\bar{n}(\eta) = N p(\eta) / (4 \pi \eta^2)$. Because the
radial distance cannot be measured directly, this function is unknown and must
be fit for in the analysis.

@@

The principle observable which we would like to consider is the correlations
between the dispersion measures of observed fast radio bursts as a function of
their angular separation.  For observed FRB events labelled by the indices $i$
and $j$, and assuming $i \neq j$, we define the dispersion measure angular
correlation function as
\begin{equation}
    w(\hat{n}_i\cdot\hat{n}_j) \equiv \langle \del D_i \del D_j \rangle.
\end{equation}
A more convenient quantity to work with is the corresponding angular
power spectrum
\begin{equation}
w(\hat{n}_i \cdot\hat{n}_j) = \sum_l \frac{2l+1}{4 \pi} C_l
P_l(\hat{n}_i\cdot\hat{n}_j).
\end{equation}

We have identified two effects that dominate the angular
correlations and which depend on the large-scale structure of the Universe and
thus cosmology.  The first effect is that the signals from two FRB events may
propagate through shared large-scale structure along the line of sight.
Thus correlations
in the election over densities along the line of sight will correlate
$\Dde$.  The second effect is that due to the intrinsic clustering of source
events, a pair of FRBs that are nearby in the angular coordinates are more
likely to also be nearby in the line of sight coordinate, thus causing
correlations in $\Dne$.  In the remainder of this section we will calculate the
contributions from each of these effects.

We assume that $D^{\rm src}_i$ are uncorrelated. Again this may not
be strictly true due to clustering of FRB sources and environmental effects
and this may bias results.  \red{I don't thing the following matter for $i \neq
j$} In addition $\Dsrc$ and $\Dne$ may be correlated due
to redshift evolution of source events. Finally, redshift evolution of the
source events and their environments could correlate $\Dsrc$ with $\Dde$. Here
we do not attempt to quantify these correlations.

@@

\red{I think this is wrong. It is missing factors of $1/(1+z)$.}

This would be true if FRB's we uniformly distributed on the sky.  However their
locations correlate with LSS, and as such we get cross terms.

@@

The joint probability of observing a pair of FRB events at 3D coordinates
$\vec x_i$ and $\vec x_j$ in infinitesimal volumes $\ud^3\vec x_i$ and
$\ud^3\vec x_j$ is
\begin{equation}
    \calP(\vec x_i, \vec x_j) =
    \frac{N^2 p(\eta_i) p(\eta_j)}{(4 \pi)^2 \eta_i^2 \eta_j^2}
    \left[ 1 + b(z_i) b(z_j) \xi(\vec x_i - \vec x_j) \right]
    \ud^3\vec x_i \ud^3\vec x_j.
\end{equation}
The factors of $1/\eta^2$ are necessary to achieve the proper volume weighting,
a fact that can be seen by taking the limit of $\xi \to 0$ and integrating of
the angular coordinates. Noting that $p(\eta_i)$ has units 
\red{$({\rm MPc}\,{\rm rad}^2)^{-1}$ (conflicts with being normalized)}.

An observed event has known angular coordinates, and we are interested in
correlations in $\Dne$ which is directly related to the line of sight
coordinate.  Fixing the angular coordinates:
\begin{equation}
    \calP(\eta_i, \eta_j | \hat{n}_i, \hat{n}_i) = N(\hat{n}_i, \hat{n}_i)
    p(\eta_i)p(\eta_j) \left[ 1 + b(z_i)
    b(z_j) \xi(\vec x_i - \vec x_j) \right] \ud \eta_i \ud \eta_j,
\end{equation}
where $N(\hat{n}_i, \hat{n}_i)$ is a normalizing factor which ensures the
expression integrates to unity over $\eta_i$ and $\eta_j$. The factors of
$1/\eta^2$ have been cancelled when converting $\ud^3 \vec x = \eta^2 \ud \eta \ud^2
\vec \theta$. If the second term in brackets is small, then the normalizations
$N$ is 1 (since the distributions $p(\eta_i)$ are already normalized. 
This expression is
only valid if $i \ne j$.
Transforming to $\Dne$ as our distance measure, we obtain
\begin{align}
\calP(\Dne_i, \Dne_j | \hat{n}_i, \hat{n}_i)
    &= p(\eta_i)p(\eta_j) \left[ 1 + b(z_i)
        b(z_j) \xi(\vec x_i - \vec x_j) \right]
        \frac{\ud \eta_i}{\ud \Dne_i} \frac{\ud \eta_j}{\ud \Dne_j}
        \ud \Dne_i\ud \Dne_j
        \\
    &= p(\eta_i)p(\eta_j) \left[ 1 + b(z_i)
        b(z_j) \xi(\vec x_i - \vec x_j) \right]
        \frac{\ud \Dne_i}{1 + z_i}\frac{\ud \Dne_j}{1 + z_j}
\end{align}
Finally, taking the second moment and subtracting off the product of the first
moments:
\begin{equation}
\langle \Dne_i \Dne_j \rangle
    - \langle \Dne_i \rangle \langle \Dne_j \rangle =
    \int \Dne_i \Dne_j
    \frac{p(\eta_i)}{1 + z_i}\frac{p(\eta_j)}{1 + z_j}
    b(z_i) b(z_j) \xi(\vec x_i - \vec x_j)
        \ud \Dne_i\ud \Dne_j
\end{equation}
And thus we have:
\begin{equation}
    w^{\bar{n}_e \bar{n}_e}(\hat{n}_i\cdot\hat{n}_j)
        =\int \Dne_i \Dne_j
    \frac{p(\eta_i)}{1 + z_i}\frac{p(\eta_j)}{1 + z_j}
    b(z_i) b(z_j) \xi(\eta_i \hat{n}_i - \eta_j \hat{n}_j)
        \ud \Dne_i\ud \Dne_j
\end{equation}
Where $\eta$ and $z$ are understood to be functions of $\Dne$.

@@

Here we sketch how the dispersion measure angular power-spectrum, $C_l$
could be estimated from a sample of observed dispersion-measures.

@@

We see that the off diagonal terms depend on the angular power-spectrum of
interest while the diagonal is contaminated by the unknown parameter
$\sigma_{\rm src}^2$. This is again in analogue to weak lensing where the
correlations between galaxy ellipticity are caused primarily by cosmic lensing shear
while an individual galaxies ellipticity is dominated the galaxy's intrinsic
ellipticity. To remove the information in the diagonal, we define a
modified covariance matrix derivative such that:
\begin{equation}
    \tilde{C}_{ij,\alpha} \equiv
\begin{cases}
    C_{ij,\alpha} & \text{if } i \neq j,\\
0 & \text{if } i = j
\end{cases}.
\end{equation}
